#Reproducible research Project 1
John Huberty

December 10, 2014

###Load necessary packages
```{r warning=FALSE, message=FALSE}
library(lubridate)
library(plyr)
library(ggplot2)
library(data.table)
library(lattice)
```

###Loading and preprocess the data

1. Load the data from activity.csv file
```{r}
setwd("C:/Users/jhuberty/Desktop/Coursera_data_science/6 Reproducible research")
activity <- read.csv("project 1/activity.csv")
```
2. Process/transform the data into a format suitable for analysis
```{r}
nrowactivity <- nrow(activity)
activity$index <- c(1:nrowactivity) 
testdate <- activity[1,2]
index <- 1

###Loop to create index that matches 5 minute time period to 24 hour clock.
```{r}
for(i in 1:nrowactivity) {
        if(activity[i,2] == testdate) {
                activity[i,4] <- index
                index <- index + 1
                i <- i+1                
        } else if(activity[i,2] != testdate) {
                testdate <- activity[i,2]
                index <- 1
                activity[i,4] <- index
                index <- index + 1
                i <- i + 1
                
        }
}


activity$hours <- (((activity$index  - 1)*5) %/% 60)      
activity$minutes <- ((((activity$index  - 1)*5) / 60)-(((activity$index  - 1)*5) %/% 60))*60 
activitynona <- na.omit(activity)
```

###What is the mean total number of steps taken per day?
```{r}
stepsday <- ddply(activitynona, "date", summarize, steps=sum(steps))
meanstepsday <- mean(stepsday$steps)
medianstepsday <- median(stepsday$steps)
```

1. Make a histogram of the total number of steps taken each day
```{r}
hist(stepsday$steps)
```

2. Calculate and report the mean and median total number of steps taken per day
       
       - The mean steps taken per day is `r meanstepsday`. 
       
       - The median steps taken per day is `r medianstepsday`.

###What is the average daily activity pattern? 

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and 
the average number of steps taken, averaged across all days (y-axis)

```{r}
stepsdayindex <- ddply(activitynona, "index", summarize, steps=mean(steps))
qplot(index, steps, data = stepsdayindex, geom = "line", ylab("Average number of step"), xlab("5 minute time interval"))
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}
stepsdayindexmax <- stepsdayindex[which.max(stepsdayindex$steps),]
stepsdayindexmax[1]
```

        - The `r stepsdayindexmax[1]`th 5 minute interval, on average across all the days in the dataset, contains the maximum number of steps. 

###Input missing values

1. Calculate and report the total number of missing values in the dataset.

```{r}
activity2 <- activity
nacount <- complete.cases(activity2)
nacountsum <- sum(!nacount)
nrowactivity <- nrow(activity2)
```
       - The total number of missing values in the dataset is `r nacountsum`.

2. Devise a strategy for filling in missing values in the dataset.
        - Strategy will be to fill in the data with the mean for the interval
        
3. Create a dataset equalt to the original data set but with the missing data filled in.
```{r}
for(i in 1:nrowactivity) {
        if(nacount[i] == "TRUE") {
           i <- i+1
        } else if(nacount[i] == "FALSE") {
           #get the interval number from the activity2 data frame where there is an na value
           naindex <- activity2[i,4]
           #get the row number of the steps_day_interval data frame that corresponds to the interval above
           rownumberreplace <- which(stepsdayindex$index == naindex)
           #use the row number found to determine the value to replace the na value with
           nastepsreplace <- stepsdayindex[rownumberreplace,2]
           #replace na steps value with mean for the interval
           activity2[i,1] <- nastepsreplace 
           #iterate to next row
           i <- i+1
        } else {
                i <- i+1
        }
}
```

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
stepsday2 <- ddply(activity2, "index", summarize, steps=sum(steps))
hist(stepsday2$steps)
meanstepsday2 <- mean(stepsday2$steps)
medianstepsday2 <- median(stepsday2$steps)
```

        - The mean steps taken per day is `r meanstepsday2`. 
       
        - The median steps taken per day is `r medianstepsday2`.

        - The values do differ from the estimates from the first part of the assignment. The impact of imputing missing data on the estimates significantly increases the median and mean steps per day. This is most likely due to attributing steps to periods of the day where none actually occured. 

### Are there differences in activity patterns between weekdays and weekends? 

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}
Sys.setenv(TZ='GMT')
activity2$date <- ymd(activity2$date)
activity2$day <- weekdays(activity2$date)
activity2$dayend <- activity2$day

# convert days to either weekend or weekday
i = 1
for(i in 1:nrowactivity) {
        if(activity2[i,7] == "Saturday" || activity2[i,7] == "Sunday") {
                activity2[i,8] <- "weekend"
                i <- i+1
        } else {
                activity2[i,8] <- "weekday"
                i <- i+1
        }
}
```
2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r}
stepsdayend <- ddply(activity2, c("dayend","index"), summarize, steps=mean(steps))
xyplot(steps ~ index | dayend, data = stepsdayend, layout = c(1,2), type = "l")
```








